---
title: "Cooccurrence"
author: "Daniel Sultanov"
date: "7/20/2022"
output: html_document
---

libraries
```{r}
library("circlize")
library("dplyr")
library("tidyverse")
```

data
```{r}
isolates <- read.table("~/Desktop/work/database/updated_tables/Variants.txt", sep = "\t", header = T)

acc_isolate <- read.table("~/Desktop/work/database/source/acc_isolate_relation.txt",
                          "\t",
                          header=T)
```

#Cooccurrence in all isolates together
```{r}
#get list of variants
variants <- unique(select(isolates,REFERENCE,POSITION,ISOLATE))
#index variants ("INDEX_type"). This will be used to identify and reference them later (don't use rownames as thewy come from original table)
variants$INDEX_type<-seq(1,nrow(variants),by=1)

##make cooccurrence matrix
#assign variant IDs (variants$INDEX_type) to isolates (ACCESSION)
accession_var_relation<-merge(isolates, variants, by=c("POSITION", "REFERENCE", "ISOLATE"))
accession_var_relation_2<-select(accession_var_relation, ACCESSION, INDEX_type)
#get variants that *occur* (but not necessarily co-occur!) in the same isolate and store it in matrix
#this matrix only counts instances when *both* variant A AND B are present
cooc_m<-crossprod(table(accession_var_relation_2))
#since the diagonal is the same variant with itself, assign zeros
diag(cooc_m)<-0
#triangularize
cooc_m[upper.tri(cooc_m,diag=T)]<-0
```

Calculate distances between iVFs for each pair using Eucledian distances. Distance==abs values VS correlation == strength of their relationship. Also need to take into consideration A NOT B AFs and B NOT A iVFs (if they are both 0's in the rest of the cases euclid dist will not be much affected).

Calculate for the whole dataset where the same variant pairs are found in AT LEAST "thld" different isolates and where mean(iVFs) >= 0.05
```{r}
#allocate empty matrix to store results
euclid_dist<-matrix(nrow=nrow(variants),ncol=nrow(variants))
#thld - more than this amount of isolates must have the same pair (i.e. thld 2 == at least 3 isolates)
#if (mean(s$AF)>0.05) - if all AFs in the comparison <0.05 - do not consider
accession_var_relation_3<-select(accession_var_relation, ACCESSION, VF, INDEX_type)

#this is for table
thld<-2
for (j in 1:nrow(cooc_m)){
  for (k in 1:ncol(cooc_m)){
    if (cooc_m[j,k]>thld){
    s<-subset(accession_var_relation_3, INDEX_type==j | INDEX_type==k)
    if (mean(s$VF)>0.05){
      euclid_dist[j,k]<-dist((spread(s, ACCESSION, VF, fill=0))[,-1])
      }else{euclid_dist[j,k]<-NA}
    }else{euclid_dist[j,k]<-NA}
  }
}
#get coordinates of those where d<0.5
euclid_0.5<-which(euclid_dist<0.5,arr.ind = T)
head(euclid_0.5)
```
Plot
```{r}
#get x-values for variant pairs
#y-vals will be just to fit it in the plot
pair_POS<-data.frame(x1<-rep(0,nrow(euclid_0.5)),x2<-rep(0,nrow(euclid_0.5)),y1<-rep(0,nrow(euclid_0.5)),y2<-rep(0,nrow(euclid_0.5)))
colnames(pair_POS)<-c("x1","x2","y1","y2")
for (i in 1:nrow(pair_POS)){
  pair_POS[i,1]<-variants[euclid_0.5[i,1],2]
  pair_POS[i,2]<-variants[euclid_0.5[i,2],2]
  pair_POS[i,3]<-nrow(pair_POS)-i #this is to assign decreasing y-vals for linear plotting
  pair_POS[i,4]<-nrow(pair_POS)-i 
}
#NB: i is also Im part of complex numbers; be careful when use in operations 
#(i.e. DO NOT forget 10*i instead of 10i when multiplying numbers)

#for indentical x1 add identical y-coord
pair_POS_y<-(unique(pair_POS$x1))
#assign new y-vals NB: they will start from 1 to END (not with hight to low like in loop above)
pair_POS$y11<-rep(0,nrow(pair_POS))
pair_POS$y22<-rep(0,nrow(pair_POS)) #do it twice so vectors are the same length when plotting
for (i in 1:nrow(pair_POS)){
  pair_POS[i,5]<-which(pair_POS_y==pair_POS[i,1])
}
for (i in 1:nrow(pair_POS)){
  pair_POS[i,6]<-which(pair_POS_y==pair_POS[i,1])
}

#for circos plot at least one sector must be assigned
#sectors are factors here
pair_circ = data.frame(sectors = rep("a",length(pair_POS$x1)),x1 = pair_POS$x1, x2 = pair_POS$x2, y = rep(0.5,nrow(pair_POS)))
#assign colors: if intra-region cooc - blue; inter- red; between rRNAs - pink
col_pairs<-rep("a",nrow(pair_circ))
for (i in 1:nrow(pair_circ)){
  if (pair_circ[i,2] %in% c(1:1243) & pair_circ[i,3] %in% c(1:1243)){
    col_pairs[i]<-"blue"
  }else if (pair_circ[i,2] %in% c(1244:1364) & pair_circ[i,3] %in% c(1244:1364)){
      col_pairs[i]<-"blue"
  }else if (pair_circ[i,2] %in% c(1365:2279) & pair_circ[i,3] %in% c(1365:2279)){
    col_pairs[i]<-"blue"
  }else if (pair_circ[i,2] %in% c(2280:2490) & pair_circ[i,3] %in% c(2280:2490)){
    col_pairs[i]<-"blue"
  }else if (pair_circ[i,2] %in% c(2491:5886) & pair_circ[i,3] %in% c(2491:5886)){
    col_pairs[i]<-"blue"
  }else if (pair_circ[i,2] %in% c(5887:6118) & pair_circ[i,3] %in% c(5887:6118)){
    col_pairs[i]<-"blue"
  }else if (pair_circ[i,2] %in% c(6119:6276) & pair_circ[i,3] %in% c(6119:6276)){
    col_pairs[i]<-"blue"
  }else if (pair_circ[i,2] %in% c(6277:6637) & pair_circ[i,3] %in% c(6277:6637)){
    col_pairs[i]<-"blue"
  }else if (pair_circ[i,2] %in% c(6638:8437) & pair_circ[i,3] %in% c(6638:8437)){
    col_pairs[i]<-"blue"
  }else if (pair_circ[i,2] %in% c(8438:9099) & pair_circ[i,3] %in% c(8438:9099)){
    col_pairs[i]<-"blue"
  }else {col_pairs[i]<-"#FF69B4"}
  }

#pdf(file = "~/Desktop/work/rDNA_project/all_isolates/cooccurrence/circ_par_OR_thld_2.pdf",
#    width = 4,
#    height = 4)

circos.par("track.height" = 0.08) #each track is 10% of the circle
circos.initialize(pair_circ$sectors, xlim=c(11,9099)) #initialize first
circos.track(pair_circ$sectors, ylim=c(0,1)) #treat is as R "plot" function; initialize second

#draw sectors = coding/reg regions
#param are in DEGREES
#change rou1 and 2 according to circos.par("track.height")
nt1<-360/(9099-11) #1 nt of rDNA equals to degress on a circle
#since clock-wise, degrees are neg
draw.sector(0, -1243*nt1, rou1 = 0.99, rou2 = 0.91, clock.wise = T, border=NA ,col = "lightgrey") #NTS2-1
draw.sector(-1244*nt1, -1364*nt1, rou1 = 0.99, rou2 = 0.91, clock.wise = T, border=NA ,col = "lightgreen") #5S
draw.sector(-1365*nt1, -2279*nt1, rou1 = 0.99, rou2 = 0.91, clock.wise = T, border=NA ,col = "lightgrey") #NTS1-2
draw.sector(-2280*nt1, -2490*nt1, rou1 = 0.99, rou2 = 0.91, clock.wise = T, col = "orange") #ETS2-1; with borders
draw.sector(-2491*nt1, -5886*nt1, rou1 = 0.99, rou2 = 0.91, clock.wise = T, border=NA ,col = "lightgreen") #25S
draw.sector(-5887*nt1, -6118*nt1, rou1 = 0.99, rou2 = 0.91, clock.wise = T, border=NA ,col = "orange") #ITS2-1
draw.sector(-6119*nt1, -6276*nt1, rou1 = 0.99, rou2 = 0.91, clock.wise = T, border=NA ,col = "lightgreen") #5.8S
draw.sector(-6277*nt1, -6637*nt1, rou1 = 0.99, rou2 = 0.91, clock.wise = T, border=NA ,col = "orange") #ITS1-1
draw.sector(-6638*nt1, -8437*nt1, rou1 = 0.99, rou2 = 0.91, clock.wise = T, border=NA ,col = "lightgreen") #18S
draw.sector(-8438*nt1, -359, rou1 = 0.99, rou2 = 0.91, clock.wise = T, border=NA ,col = "orange") #ETS1-1

#both here are blue (can do in one line but col's can be changed pair-wise)
circos.trackPoints(pair_circ$sectors, pair_circ$x1, pair_circ$y, pch=16, cex=0.7, col="red")
circos.trackPoints(pair_circ$sectors, pair_circ$x2, pair_circ$y, pch=16, cex=0.7, col="red")
for (i in 1:nrow(pair_circ)){
  circos.link("a",pair_circ[i,2],"a",pair_circ[i,3], h.ratio = 0.5, col=col_pairs[i])
}
```
Write into a table
```{r}
candidates<-cbind(euclid_0.5,pair_POS[,1:2])
candidates$size<-rep(0,nrow(candidates))#sample size
colnames(candidates)<-c("var1","var2","x1","x2","size") #rename "row" and "col" from the table
#var1 corresponds to row names in cooc_m, var2 - colnames in cooc_m
head(candidates)
#add sample size to the table; here size is the number of NON-ZERO pairs! (AF A & B != 0)
#this differs from (spread(subset(accession_var_relation_3, INDEX_type=="A_val" | INDEX_type=="B_val"), ACCESSION, AF, fill=0))[,-1] , which will give you higer value bc missing vals for A or B will be counted
for (i in 1:nrow(candidates)){
  candidates[i,5]<-cooc_m[candidates[i,1],candidates[i,2]]
}
candidates

##create table
#!!NOTE!!: MUST be the same parameters as for plotting - e.g. here eucledian dist threshold < 0.5, mean < 0.5 and thld == 2
cooc_isolates<-data.frame(c("a"), c(0), c(0), c(0), c(0))
colnames(cooc_isolates)<-c("ACCESSION", "VF.x", "INDEX_type.x", "VF.y", "INDEX_type.y")

for (j in 1:nrow(cooc_m)){
  for (k in 1:ncol(cooc_m)){
    if (cooc_m[j,k]>thld){
      s<-subset(accession_var_relation_3, INDEX_type==j | INDEX_type==k)
      if (mean(s$VF)>0.05){
        if(dist((spread(s, ACCESSION, VF, fill=0))[,-1]) < 0.5){
          u <- s
          v <- subset(u, INDEX_type == j)
          w <- subset(u, INDEX_type == k)
          q <- merge(v, w, by = "ACCESSION")
          cooc_isolates<-rbind(cooc_isolates,q)}
      }
    }
  }
}
cooc_isolates <- cooc_isolates[-1,]

#add names
isolates_coocurrence <- left_join(cooc_isolates, acc_isolate, by = "ACCESSION")
isolates_coocurrence <- isolates_coocurrence[,1:6]
colnames(isolates_coocurrence) <- c("ACCESSION", "VF.x", "INDEX_type.x", "VF.y", "INDEX_type.y", "Standardized.name")

#explicitly add variants; .x = one variant from a pair, .y = its partner in the pair
#do it column by column

#!!!!!!!
#changing colnames
#!!!!!!!
colnames(variants) <- c("REFERENCE", "POSITION", "ISOLATE", "INDEX_type.x")
cooccurrence_table <- left_join(isolates_coocurrence, variants, by = "INDEX_type.x")
colnames(variants) <- c("REFERENCE", "POSITION", "ISOLATE", "INDEX_type.y")
cooccurrence_table <- left_join(cooccurrence_table, variants, by = "INDEX_type.y")
cooccurrence_table <- select(cooccurrence_table, Standardized.name, ACCESSION, POSITION.x, ISOLATE.x, VF.x, POSITION.y, ISOLATE.y, VF.y)

#filter out instances of cooccurred variants in the same position
#example: 
#CTTTGGAA
#CGTTTGGAA are Indels at a same position but can be recognized as cooccurred variants
cooccurrence_table <- subset(cooccurrence_table, !(POSITION.x == POSITION.y))

#change colnames one more time
colnames(cooccurrence_table) <- c("NAME", "ACCESSION", "POSITION_1", "VARIANT_1", "VF_1", "POSITION_2", "VARIANT_2", "VF_2")
```



